An Empirical Analysis of the Internet Engineering Task Force with Computational Methods

### 2024-09-22 paper.tex
Add content from ICWSM 2024 paper, section 4

\newlength{\figureWidthOneColumn}
\setlength{\figureWidthOneColumn}{0.40\textwidth}

@@

@@

@@

@@

>>>>

\newlength{\figureWidthOneColumn}
\setlength{\figureWidthOneColumn}{0.4\textwidth}

\newlength{\figureWidthTwoColumn}
\setlength{\figureWidthTwoColumn}{0.8\textwidth}

@@

For \textbf{RQ1}, to determine the steepness or diffusion of the hierarchy,
the proportion of nodes and activity are determined in each hierarchy
level. Figure~\ref{fig:activity_proportions}a shows the number of active
nodes in each level. The bottom level, RPs, are the largest contingent
consisting of 90--86\% of active participants; this proportion is
decreasing over time. The next level is WGCs which manage the WGs and are
6--10\% of active participants; this proportion is rising over time. The
top level is ADs which manage many WGs within an area and consist of less
than 1\% of the total active population of the IETF; this stays constant
over time. Therefore, the WGC level is becoming more diffused over time as
there are about 15 RPs for every one WGC in the organisation in 2013 versus
8 RPs per WGC in 2021.

The proportion of activity of nodes in the last year is also plotted and
split into hierarchy level in Figure~\ref{fig:activity_proportions}b. It is
clear that individuals at higher levels contribute a disproportionate share
to communication and this share is increasing over time, which is
consistent with findings in \cite{khare2022web}. Therefore, less of a
proportion of communication is flowing from the RPs at the same time as the
number of WGC roles is increasing. This further suggests that the IETF
hierarchy is becoming more diffused, in terms of mailing list
communication, which may make cooperation between hierarchy levels more
difficult.

\begin{figure}[t]
  \centering
  \small{a)} \\
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/active_node_proportions_year_sliding_window.png} \\
  \small{b)} \\
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/activity_proportions_year_sliding_window.png}
  \caption{
    Proportions of RPs, WGCs and ADs by a) number of active individuals and
    b) activity (number of emails sent to mailing lists). 
  }
  \label{fig:activity_proportions}
\end{figure}

Figure~\ref{fig:wgc_over_time}a shows the amount of nodes that inhabit the
WGC hierarchy level over our time period. There is a $35\%$ rise in the
amount of WGC roles whereas the amount of individuals that fill the WGC
roles remains mostly constant.

Figure~\ref{fig:wgc_over_time}b contains two estimates of the number of WGs
over time. The estimated number of WGs is likely the most accurate in the
middle 50\% of both plotted lines. Their discrepancy is the least (at most
10 WGs) from late 2014 to early 2019. The start of the COVID-19 pandemic
may also affect mailing list activity in 2020, and similar trends are
visible in the data in \cite{mcquistin2021characterising}. A full
explanation on how the data is estimated can be found in
section~\ref{sec:WG_activity}.

Figure~\ref{fig:wgc_over_time}c shows how many WGCs exist per WG, using
both estimates of the WG amount from Figure~\ref{fig:wgc_over_time}b. It is
clear there is a slow a rise in WGCs per WG over time. The ratio rises from
between 1.7-1.9 to 2.0-2.2 within the middle 50\%, late 2014 to early
2019.This perhaps represents a growing understanding in the IETF that it is
desirable to have two or more chairs per WG in case of
illness/unavailability or to avoid conflicts of interest. Therefore, this
plot again shows that the hierarchy is becoming more diffused, even if the
raw amount of individuals in the WGC roles has remained constant. The
multiple roles per individual may result in even more difficulty of RPs to
use their voice.

An AD version of Figure~\ref{fig:wgc_over_time} is available, but it is
omitted to save space as all three plots stay mostly constant throughout
the period. There are about 15 individual ADs split amongst 15 roles in 7-8
areas with about 2 ADs per Area. The set of IETF Areas is broadly fixed,
while WGs are created and closed relatively frequently.

\begin{figure}[t]
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/over_time.png}
  \caption{
    Overview of WGC numbers and WG numbers over time. a) shows the number
    of WGC roles (some individuals hold multiple WGC roles) and the number
    of individuals who are WGCs b) shows two different estimates of the
    number of WGs (see section~\ref{sec:method}) and c) shows the mean
    number of WGCs per WG. Notice the non-zero-based y-axis exaggerates the
    variation in these figures.
  }
  \label{fig:wgc_over_time}
\end{figure}

An AD version of Figure~\ref{fig:wgc_over_time} is available, but it is
omitted to save space as all three plots stay mostly constant throughout
the period. There are about 15 individual ADs split amongst 15 roles in 7-8
areas with about 2 ADs per Area. The set of IETF Areas is broadly fixed,
while WGs are created and closed relatively frequently.

Table~\ref{tab:WGC_email} shows the emails sent and received for WGC and
ADs on mailing lists in the year before they first take that role, and the
year after. WGCs and ADs are both active in using the mailing lists, ADs
much more so. However, they don't become more active to any major degree
after taking the roles. This may suggest that the individuals who take on
higher hierarchy level roles are already contributing at a higher level.

\begin{table}[!tbp]
\centering
    \begin{tabular}{c||c|c}
        Status & Sent & Received \\
         \hline \hline 
        Before WGC & $67.26\pm99.80$ & $70.11\pm103.37$ \\
        \hline
        After WGC & $77.20\pm102.85$ & $85.90\pm107.38$ \\
        \hline
        Before AD & $218.84\pm226.92$ & $230.50\pm214.75$ \\
        \hline
        After AD & $209.21\pm186.46$ & $219.16\pm189.85$ \\
    \end{tabular}
        \caption{Mean number of emails received for one year before/after becoming a WGC (AD) for the first time. The error is one standard deviation.}
    \label{tab:WGC_email}
\end{table}

\pb{Summary}
In answer to RQ1 the picture is mixed, the number of individuals with WGC
roles has remained broadly constant but the number of WGC roles has
increased, indicating that individuals willing to become WGC have taken on
more such roles. The number of the higher level AD roles has, by design,
remained broadly constant over the time studied. The proportion of active
individuals with WGC roles has grown from 6\% to 10\% and the proportion of
the conversation taken by those roles has grown from 25\% to 35\% over the
period studied. Therefore the hierarchy has become more diffused, which may
mean RPs experience more difficulty in expressing their voice.

@@

To quantify the effect of being in each hierarchy level has on network wide
communication for \textbf{RQ2}, we calculate the proportion of three node
motifs, lasting at most a month, in which each node participates over a
year period, see section~\ref{sec:motif}. These motifs are counted,
categorised into the three hierarchy levels and then proportions are taken
of each motif type.

Figure~\ref{fig:motif_proportions} shows the proportions of each level for
three node ``Inward Star", ``Outward Star", ``Mixed Star", and
``Triangles". The Inward and Mixed Star motifs show about a $4\%$ and
$10\%$ increase in proportion for higher levels in the hierarchy versus for
RPs. However, about a $10\%$ increase for RPs is seen for Triangles versus
WGCs and ADs, and Outward Star proportions remain similar for all levels of
the hierarchy.

The larger proportion for WGCs and ADs of Inward and Mixed Star motifs
whilst the Outward Star motifs remain similar for all levels suggests that
higher levels receive more direct communication. The higher proportion of
RP Triangle motifs suggests their discussion is more of a group activity,
whereas WGCs and ADs have a larger proportion of one-on-one conversations.
Therefore, we interpret the general communication patterns within the IETF
as a discussion amongst RPs interspersed with questions sent to and
announcements from WGCs and ADs. This indicates some collaboration between
layers and confidence of some RPs' to voice their opinions upwards.

\begin{figure*}[t]
  \centering
  \includegraphics[width=\figureWidthTwoColumn]{figures-prev/icwsm-2024/motif_strata_comparison.png}
  \caption{
    Communication patterns analysed by three edge temporal motifs for RPs,
  }
  \label{fig:motif_proportions}
\end{figure*}

Figure~\ref{fig:origin_emails} seems to corroborate this. We calculate the
proportion of mailing list threads which nodes in each hierarchy level
originate, using a year window, pushed forward by a month each calculation.
We see that WGCs and ADs send a disproportionate number of originating
emails to the mailing lists. WGCs send 30--50\% whereas they are 6--10\% of
active individuals, and ADs 5--10\% versus 0.5\%.

Therefore, the larger proportion of inward motifs for higher levels is in
part due to WGCs and ADs disproportionately originating email threads.
Other IETF participants will then reply to the thread, boosting the
proportion of inward motifs WGCs and ADs appear in. This may suggest the
higher levels are good ``condensers of communication" as they receive more
than they send.

In answer to the RQ, combining the interpretations of the two plots, the
discussion seems to happen in the following way: The ADs or WGCs are
commonly the originators of discussion threads. Where conversation includes
a WGC or AD the conversation tends to be directed around them and usually
inward towards the WGC or AD. The RPs are more likely to engage in
discussions amongst themselves (triangular communication patterns) when
higher level individuals are not present.

\begin{figure}[t]
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/Origin_email_proportions.png}
  \caption{
    Proportion of ``origin" (non reply) emails sent by RPs, WGCs and ADs,
    this should be interpreted in conjunction with
    figure~\ref{fig:activity_proportions} showing the proportion of active
    individuals in each class.
  }
  \label{fig:origin_emails}
\end{figure}

@@

For \textbf{RQ3}, we determine the direction communication tends to flow
through the hierarchy by looking at inbound and outbound communication
flows. The network's edges are categorised based on their source and target
node's hierarchy level. For each of the inter-level communication
combinations the proportion of communication in the last year going in the
``upward" hierarchy direction is calculated. For instance, if there are six
emails from RPs to WGCs and twelve from WGCs to RPs then the proportion
upward is $\frac{6}{6+12}=\frac{1}{3}$.

In Figure~\ref{fig:inter_strata}, these proportions are plotted, and the
year long window is again pushed forward by one month to increase the
resolution of changes. A proportion of higher than $0.5$ shows more
communication flowing up the hierarchy than down. All three levels are
mostly upward in direction, with the lowest two levels (RP$\rightarrow$
WGC) having the largest skew in the proportion. The other two lines
(RP$\rightarrow$AD and WGC$\rightarrow$AD) show similar periodicity, with
the WGC$\rightarrow$AD line dropping as far below $0.5$ as above showing a
large shift in inter-level communication patterns higher up the hierarchy.

The direct measurement of message flow between the hierarchy levels seems
to bolster our interpretation that the cooperative design of the IETF is
working. This analysis shows that RPs are communicating much more
preferentially upwards in the hierarchy, despite the decrease in overall
share of RP activity. The increase within the WGC level does not have any
noticeable effect on their level of communication, if anything there is a
negative relationship. This again suggests both WGCs and ADs perform a kind
of ``condenser of communication" or ``facilitator" role. In other words,
this analysis suggests the IETF has a ``bottom-up" communication style,
where the higher levels encourage the lower levels' voice.  This is the
desired pattern of communication for a volunteer organisation that is
looking to grow and encourage participants to engage and gradually take on
leadership roles.

Overall for RQ3 then, the pattern observed was communication flowing up the
hierarchy and this corresponds to the answer found in RQ2. WGCs and ADs
were more likely to receive communication with individuals than send
communication to individuals whereas both were more likely than RPs to send
out messages to the list in general.

\begin{figure}[t]
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/up_hierarchy_communication_ratios.png}
  \caption{
    Proportion of communications that are ``up" the hierarchy for different
    groups. A proportion above 0.5 indicates that the majority of
    communications are ``upward".
  }
  \label{fig:inter_strata}
\end{figure}

@@

Finally, to determine how individuals and their neighbours affect each
other's activity levels in subsequent time periods, for \textbf{RQ4}, we
take correlations of degree between different time windows using the
Mobility Taxonomy~\cite{barnes2023measuring} aspects called
\textit{Mobility}, \textit{Neighbour Mobility}, \textit{Philanthropy} and
\textit{Community} explained in the previous section. In
Figure~\ref{fig:mob_tax}, the time window chosen is one year which is split
into two snapshots graphs of six months, the time window then moves forward
one month and the process is repeated. The mid point of the year time
window is plotted. WGCs and RPs are plotted but the small number of ADs
mean that the correlations are too noisy to show meaningful effects.

\begin{figure*}[t]
  \centering
  \includegraphics[width=\figureWidthTwoColumn]{figures-prev/icwsm-2024/hierarchy_status_comparison.png}
  \caption{
    Connection between roles and activity for RP and WGC viewed in terms of
    a) Mobility b) Neighbour Mobility c) Philanthropy and d) Community over
    time. See section~\ref{subsec:mob_tax} for definitions of these terms.
  }
  \label{fig:mob_tax}
\end{figure*}

The \textit{Mobility} of both RPs and WGCs is plotted in
Figure~\ref{fig:mob_tax}a. The high amount of correlation between the
degree of an individual in the first half of the time window with the
second half shows that all users have a tendency to become more active in
subsequent time periods if they are highly active in the previous period,
but this is more pronounced in WGCs. This is found to follow a period of
about one year for both RPs and WGC, suggesting that a person is likely to
be as active as there were a year ago. The \textit{Neighbour Mobility}
plot, Figure~\ref{fig:mob_tax}b, shows a similar tendency for
neighbourhoods, but with less of a split.

The \textit{Philanthropy} plot, Figure~\ref{fig:mob_tax}c, shows a higher
correlation for WGCs than RPs between an individual's degree in the first
snapshot with and their ND in the second. This means that individuals who
directly interact with WGCs become more active on mailing lists
subsequently. The \textit{Community} plot, Figure~\ref{fig:mob_tax}d, shows
the correlation between the ND in the first snapshot with the degree in the
second.  The higher correlation for WGCs suggests they benefit from
interactions with highly active neighbours by themselves becoming more
active in subsequent time periods. These trends for \textit{Philanthropy}
and \textit{Community} show a reversal in early 2016 that coincides with a
period where the mean number of WGCs per WG and the number of WGC roles was
rapidly rising, see figure~\ref{fig:wgc_over_time}. A merging of the
Applications area and Real-time Applications \& Infrastructure area in May
2015 may be a cause of this reversal of correlation, as the assignment of
WGs to areas was changed which may cause more non-reciprocal WGC
communication than normal.

Increased \textit{Mobility} for WGCs suggests that they gain advantage from
their role with their activity leading to similar communication activity in
the future. Moreover, they experience an increased effect of
\textit{Philanthropy} and \textit{Community}. The former suggests that if a
WGC is engaged in a higher amount of discussion, then in the future the
individuals who they were communicating with will also be highly active.
The latter suggests that when WGCs are surrounded by people who are active
in discussion, they are likely to be more active themselves in the future
too. All of this has less of an effect for RPs who show a lower correlation
in terms of all three aspects. Our interpretation of this is that active
WGCs are ``facilitating" discussion in the WG mailing lists; both having
their activity boosted by their neighbours and, in turn, boosting the
activities of their neighbours. This aligns well with the hypothesis that
the cooperative and voluntary design of the IETF lends itself well to
higher levels encouraging those lower to engage in discussion.

Overall the answer to RQ4 is that the WGCs have a positive effect on their
immediate contacts and those individuals that directly discuss with WGCs
are more likely to engage more in discussion in subsequent time periods.
Conversely though the WGCs who discuss topics with individuals who are
active in discussion are, themselves, more likely to engage more fully in
discussion. This points to something like a virtuous cycle of WGCs
encouraging and being encouraged by their direct contacts.

% From our ICWSM 2024 paper, Section 6:

\todo{summarise hierarchy results}
